---
title: "basic r cheatsheet"
output: html_document
---

### Base R
We probably aren't going to come back to it now that we use things like the tidyverse, but it's still good to know. However, there are still important concepts!

## Basic Arithmetic
We have data... cool! Remember the following if you want to do basic arithmetic in R...
1. R follows PEMDAS, so format your math accordingly.
2. `+` is for addition, `-` is for subtraction.
3. `*` is for multiplication, `/` is for divison.
4. `**` is for exponents.
5. You can take the `mean()` or `median()` of your data, as long as it is in an object!

### Types of Data
## Classes
These are the main five `classes` of data in R.
```{r}
my_numeric <- 42
my_integer <- 2L
my_character <- "universe"
my_logical <- FALSE
my_complex <- 2+4i
```

If you don't know the type of class your data falls under, just use the `class()` function. It will return the class type.
You can use the `is()` function to clarify what type of class type your data is -- this returns a `TRUE` or `FALSE`.
```{r}
is.integer(my_character)
```
```{r}
is.integer(my_integer)
```

The `as()` function is good for specifying your data from one class to another.
```{r}
my_integer <- as.integer(my_numeric)
```

## Missing Data
When you are missing data, R designates it as the NA. There are two ways to find out if and where NAs are in your data.
The `is.na` function is useful for a simple `TRUE` or `FALSE` about whether NAs exist in your data.
```{r}
#is.na(my_data)
```
The `anyNA` function is useful in telling you where exactly the NA occurs
```{r}
#anyNA(my_data)
```

