---
title: "coding cheat sheet"
output: 
  html_document: 
    keep_md: yes
---

## Lab 1: Getting Started
# Working directory
```{r}
getwd()
```

# Getting help
```{r}
?mean
```


# Arithmetic
```{r}
4*12 # multiplication
4/2 # division
4 + 2 # addition
4 - 2 # subtraction
8**2 # exponential
```
R follows the order of operations (PEMDAS).

# Objects
```{r}
x <- c(2, 3, 4)
```
The object 'x' should show up in your global environment.

# Statistics
```{r}
mean()
median()
sum()
```

## Lab 2: Objects, Classes, NAs, Vectors, Data Matrices
# Data types
```{r}
my_numeric <- 42
my_integer <- 2L #adding an L automatically denotes an integer
my_character <- "universe"
my_logical <- FALSE
my_complex <- 2+4i
```

# Finding data types
```{r}
class()
is.integer() # is my object an integer? returns boolean TRUE/FALSE
as.integer() # sets your object to class of choice

```

# Missing data
```{r}
is.na() # is there an `NA` in my data right now? reutnrs TRUE/FALSE
anyNA() # shows you where there is an NA in your data
```
When executing certain functions, you can exclude the NAs being calculated by writing `na.rm = T`.


