---
title: "coding cheat sheet for data frames"
output: 
  html_document: 
    keep_md: yes
---
***THINGS TO ASK LEDFORD***
1. When we get to the midterm, do you want us to answer things using primarily functions from tidyverse/janitor?
2. Would we get docked off points if we are using R Base Code?
3. Is identifying vector elements and data matrices something we will utilize anymore in the class?

### Getting Set-Up
Make sure your workspace is set up. Check your working directory.
```{r}
getwd()
```

Load your libraries, because we shouldn't rely on using R Base Code (although it's not a bad thing).
```{r}
library(tidyverse)
library(janitor)
```

Get your data imported.
```{r}
ps4_games <- readr::read_csv("data/PS4_GamesSales.csv")
xbox_one_games <- readr::read_csv("data/XboxOne_GameSales.csv")
video_game_sales <- readr::read_csv("data/Video_Games_Sales_as_at_22_Dec_2016.csv")
```
See that `<-`? We just told R to "get" the .csv containing our data and store it in the **object** called `ps4_games`.

### Dataframes
## Examining Your Data Structure: Dataframes
Summary functions are a great way to get an idea of data structure, content, and dimensions. *(The following data displays sales in the millions).*
`summary()`
```{r}
summary(ps4_games)
```
`glimpse()`
```{r}
glimpse(ps4_games)
```
`nrow()` gives number of rows.
```{r}
nrow(ps4_games)
```
`ncol()` gives number of columns.
```{r}
ncol(ps4_games)
```
`dim()` gives the dimensions.
```{r}
dim(ps4_games)
```
`names()` gives the column names.
```{r}
names(ps4_games)
```
`head()` gives the first `n` rows of data.
```{r}
head(ps4_games, n = 10)
```
`tail()` gives the last `n` rows of data.
```{r}
tail(ps4_games, n = 10)
```
`view()` is good for viewing data in a more aesthetic tab.
```{r}
view(ps4_games)
```
`table()` produces fast counts of a number of observations in a variable.
```{r}
table(ps4_games$Genre)
```


## Cleaning Your Data: Dataframes
If your data has... less than ideal column names, clean them up in two ways:

**Using the Janitor library**
```{r}
ps4_games <- janitor::clean_names(ps4_games)
names(ps4_games)
```

**Using the `rename()` function**
```{r}
names(xbox_one_games)
xbox_one_games <- rename(xbox_one_games, pos = "Pos", game = "Game", year = "Year", genre = "Genre", publisher = "Publisher", north_america = "North America", europe = "Europe", japan = "Japan", rest_of_world = "Rest of World", global = "Global")
names(xbox_one_games)
```



